Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-04-18 doc/invt.tex
editing invt.tex

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
The above two introductory sections help fix our notations,
and an expert reader can readily skip them.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses metadynamics
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time or extensive sampling limit
in Sec. \ref{sec:cmpschemes}.

@@

We will broadly call a biased simulation
targeting a flat (i.e. $p_i \equiv 1/n$)
or nearly flat\cite{
dayal2004, *trebst2004, *zhang2007, barducci2008, singh2011}
$p_i$
a flat-distribution-sampling (FDS) simulation.
%
For the biased distribution, $\pi_i$,
to converge to $p_i$,
the bias potential must satisfy
%
\begin{equation}
  V_i \to \ln p^*_i - \ln p_i + \mathrm{const.}
  .
  \label{eq:Vi_target}
\end{equation}
%
%according to Eq. \eqref{eq:pi_p_V}.
%which allows us to deduce the PMF, $-\ln p^*_i$,
%directly from the bias potential.
%
In practice,
the bias potential is adjusted to make the
observed distribution, or the normalized histogram,
match the intended one,
and once this is done,
the PMF, $-\ln p^*_i$, can be naturally
deduced from the bias potential. % $V_i$.
%
%This relation thus sets a practical aim for the convergence of FDS
%methods---the bias potential must be adjusted to make the
%observed distribution, or the normalized histogram,
%match the intended one.
%%
%Once this aim is achieved,
%the PMF, $-\ln p^*_i$, can be naturally
%deduced from the bias potential. % $V_i$.
\note{
  For example, in the context of simulated tempering (Zhang and Ma, 2007),
  $p^*_i$ is replaced by $\ln Z_i$,
  $p_i$ is the weight $\zeta_i$,
  and the bias potential, $V_i$, is $\ln (\tilde Z_i / \zeta_i)$.
  Upon convergence, we have $\ln \tilde Z_i \to \ln Z_i$
  up to a constant shift.
}

Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(estimated a priori)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, *lee1993}
and simulated tempering\cite{
marinari1992, *lyubartsev1992}.
%with $i$ being the index of energy
%and that of temperature, respectively.
%
The estimated bias potentials at bin $i$, $V_i$,
are often improved
with the normalized histogram of the sampling, $H_i$, accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_V}
allows us to correct $V_i$ as
%
\begin{equation}
  V^\mathrm{corr}_i
  =
  V_i
  +
  \ln \frac{ H_i }
           { p_i }.
  \label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
%However, for an unknown system,
%it is often difficult
%to guess a sufficiently accurate
%$V_i$ with which to start.
%%
%Thus,
%it can take many iterative rounds of simulations
%to achieve the desired flat distribution.
%%
%%Another difficulty of this approach is that
%The effectiveness of this approach
%depends on the precision of the histogram, $H_i$,
%which in turn depends on the simulation length
%that must be decided in advance.
However, this formula often requires a long
simulation period to accumulate a precise histogram, $H_i$,
and thus it disallows
continuous improvement of the bias potential
(hence that of the PMF)
as simulation lengthens.

The second type of methods,
including the WL algorithm and metadynamics,
address the above shortcomings
by continuously and incrementally updating the bias potential.
%For this reason, the second type of methods,
%referred to as the adaptive FDS methods,
%were invented.
%
These methods, refered to as the adaptive FDS methods,
will be discussed below.
%%
%%These methods incrementally update the bias potential
%%and can approximately sample the target distribution.
%%
%However,
%the on-the-fly updates break the microscopic reversibility
%of the underlying sampling mechanism.
%Further, failure to properly reduce the magnitude of
%the histogram-based increments
%would lead to roughness or errors in the final estimate of the PMF.
%%
%%Thus, for a conservative practitioner,
%%the E-FDS methods appear more appealing
%%once a sufficiently accurate bias potential is available.
%%
%To minimize this effect,
%many adaptive methods adopt a schedule to
%reduce the updating magnitude
%over the course of simulation\cite{
%marsili2006,
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%barducci2008}.
%%
%%Thus, in late stages of a long simulation,
%%the updating becomes negligible and
%%the adaptive simulation tends to
%%the corresponding equilibrium one.
%%
%%if the bias potential were converged to the exact value,
%%this adaptive FDS tends to the ideal equilibrium counterpart,
%%particular equilibrium simulation can be thought as
%%the optimal asymptotic limit of the adaptive counterpart,
%%
%For example, one can show that
%if the magnitude is controlled by
%the optimal inverse-time prescription\cite{
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%morozov2007, zhou2008,
%komura2012, *caparica2012, *caparica2014},
%the WL algorithm is equivalent to the ideal equilibrium counterpart
%in improving the precision of the bias potential,
%thus allowing for maximum efficiency
%(cf. Appendix \ref{sec:equilerr}).
%%the bias potential in the WL algorithm
%%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%%with the precision of the histogram as simulation lengthens.

@@

One can show that with a constant magnitude,
$\alpha(t) = \alpha > 0$,
the distribution collected from
an adaptive FDS simulation
is identical to $p_i$ in the long time or asymptotic limit.
%
However, as the above updates
change $V_i(t)$ continuously,
the equilibrium distribution,
Eq. \eqref{eq:pi_p_V}, no longer holds
at all times,
and the error in $V_i(t)$
depends on the updating magnitude\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, zhou2008,
  laio2005, bussi2006, poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}.
%%
%The source of the deviation is two-fold.
%%
%First, since $V_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to reduce the magnitude, $\alpha$, over time
to improve the estimate of the distribution,
$p^*_i$. %via Eq. \eqref{eq:v_def}.

@@

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
An interesting explanation\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average computed from all data points collected so far,
so that the addition of a new data point only
affects the average by its due weight, $1/t$,
which is the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
Here we will re-derive this formula
and suggest a slight modification
whereby the origin of $t$ is shifted
according to the initial error.
%
Interestingly, we may reach the same result
from optimally planning the simulation lengths in the stages
and the reduction factor, as shown in Appendix \ref{sec:wltoinvt}.

%As Eq. \eqref{eq:alpha_invt} was being developed for the WL algorithm,
%several similar ways of
%reducing the updating magnitude
%were developed for umbrella sampling and metadynamics,
%particularly, the self-healing umbrella sampling
%and well-tempered metadynamics\cite{
%  marsili2006, barducci2008, dickson2011, dama2014}.
%%
Our aim here is to find the optimal schedule, $\alpha(t)$,
for a general updating scheme.

@@

In this section,
we will derive the optimal $\alpha(t)$
for the single-bin scheme,
Eq. \eqref{eq:wl_update}.
%
To do so,
we will first express the error of
the bias potential, $V_i(t)$,
as a functional of $\alpha(t)$,
and then minimize it by variation.

\subsubsection{Error}

First, we need to define the error in the bias potential.
%
This is done by deducting two contributions from the bias potential.
%
The first contribution comes from the target value,
Eq. \eqref{eq:Vi_target}.
%
For this purpose,
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  V_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i$'s are
equivalent to the updates to $V_i$'s,
since the shift
remains a constant during the course of simulation,
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V},

@@

Note that the distribution, $\pmb \rho$,
introduced above to define the error
is independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is a flat distribution,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
  barducci2008, dama2014} ($\ln p_i = \ln {p^*_i}/(1+\Delta) + \mathrm{const.}$ for some positive $\Delta$).
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%

@@

%
For the random noise part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  =
  \sigma_{ij}(t - t'),
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function
that vanishes at large $t$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%
Below, we will assume the noise is white,
such that
$$
\sigma_{ij}(t) = \Gamma_0 \, \delta(t)
$$
for some $\Gamma_0$,
and $\delta(t)$ is the Dirac's delta function.
%
This approximation is valid if the simulation period, $T$,
is much longer than the autocorrelation time of $\sigma_{ij}(t)$.
%
For non-white noise,
the equivalent value of $\Gamma_0$ can be obtained
from the integral of the autocorrelation function,
$\int_{-\infty}^\infty \sigma_{ij}(t) \, dt$.

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also governed by white noise as is $\zeta_i(t)$,
i.e. we may write
$$
  \langle \xi_i(t) \, \xi_i(t') \rangle = \Gamma_i \, \delta(t - t'),
$$
for some $\Gamma_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

The total error
at the end of the simulation, $t = T$,
as defined in Eq. \eqref{eq:error_sum},
now depends implicitly on the function $\alpha(t)$
via $x_i(T)$.
%
Our aim is to find the $\alpha(t)$
that minimizes this error.
%
It suffices to consider
a one-variable problem that
minimizes each individual
$\left\langle x_i^2(T) \right\rangle$.

\subsubsection{\label{sec:onevar}
Optimization of a one-variable problem}

Consider the following stochastic differential equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is some white noise, such that
%
$
  \left\langle \xi(t) \, \xi(t') \right\rangle
  =
  \Gamma \, \delta(t - t')
  .
$
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 },
\label{eq:alpha_invt1}
\end{equation}
%
where $t_0$ is a constant.

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(T)
=
x(0) \, e^{-q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
u(q')
&\equiv
e^{q' - q(T)}.
\label{eq:u_def}
\end{align}
%
At the end of period $T$, we have
%
\begin{align}
  &\left\langle x^2(T) \right\rangle
  =
  \langle x^2(0) \rangle \, e^{-2 \, q(T)}
  +
  \Gamma \int_0^T
    {\dot u}^2\bigl( q(t) \bigr) \,
    \, dt.
  \label{eq:x2t_average}
\end{align}

>>>>

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
The above two introductory sections help fix our notations,
and an expert reader may wish to skip them.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses metadynamics
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix},
and present a general procedure of computing the optimal schedule
in Sec. \ref{sec:procedure}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time or extensive sampling limit
in Sec. \ref{sec:cmpschemes}.

@@

We will broadly define
a flat-distribution-sampling (FDS) simulation
as a biased simulation
targeting a flat $p_i$ (i.e. $p_i \equiv 1/n$)
or nearly flat one\cite{
dayal2004, *trebst2004, *zhang2007, barducci2008, singh2011}.
%
For the biased distribution, $\pi_i$,
to converge to the desired $p_i$,
the bias potential must satisfy
%
\begin{equation}
  V_i \to \ln p^*_i - \ln p_i + \mathrm{const.}
  .
  \label{eq:Vi_target}
\end{equation}
%
%according to Eq. \eqref{eq:pi_p_V}.
%which allows us to deduce the PMF, $-\ln p^*_i$,
%directly from the bias potential.
%
This relation thus sets a practical goal for the convergence of FDS
methods---the bias potential must be adjusted to make the
observed distribution match the intended one.
%%
Once this is done,
the PMF, $-\ln p^*_i$, can be naturally
deduced from the bias potential. % $V_i$.
\note{
  For example, in the context of simulated tempering (Zhang and Ma, 2007),
  $p^*_i$ is replaced by $\ln Z_i$,
  $p_i$ is the weight $\zeta_i$,
  and the bias potential, $V_i$, is $\ln (\tilde Z_i / \zeta_i)$.
  Upon convergence, we have $\ln \tilde Z_i \to \ln Z_i$
  up to a constant shift.
}

In terms of how one adjusts the bias potential,
there are roughly two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(estimated a priori)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, *lee1993}
and simulated tempering\cite{
marinari1992, *lyubartsev1992}.
%with $i$ being the index of energy
%and that of temperature, respectively.
%
After a sufficiently long run, the bias potential, $V_i$,
can often be improved
with the normalized histogram, $H_i$, from the simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_V}
allows us to correct $V_i$ as
%
\begin{equation}
  V^\mathrm{corr}_i
  =
  V_i
  +
  \ln \frac{ H_i }
           { p_i }.
  \label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
%However, for an unknown system,
%it is often difficult
%to guess a sufficiently accurate
%$V_i$ with which to start.
%%
%Thus,
%it can take many iterative rounds of simulations
%to achieve the desired flat distribution.
%%
%%Another difficulty of this approach is that
%The effectiveness of this approach
%depends on the precision of the histogram, $H_i$,
%which in turn depends on the simulation length
%that must be decided in advance.
However, this formula often requires a long
simulation period to accumulate a precise histogram,
and thus disallows
continuous improvement of the bias potential
%(hence that of the PMF)
as simulation lengthens.

The second type of methods,
including the WL algorithm and metadynamics,
addresses the above shortcomings
by continuously and incrementally updating the bias potential.
%For this reason, the second type of methods,
%referred to as the adaptive FDS methods,
%were invented.
%
These methods, referred to as the adaptive FDS methods,
will be discussed below.
%%
%%These methods incrementally update the bias potential
%%and can approximately sample the target distribution.
%%
%However,
%the on-the-fly updates break the microscopic reversibility
%of the underlying sampling mechanism.
%Further, failure to properly reduce the magnitude of
%the histogram-based increments
%would lead to roughness or errors in the final estimate of the PMF.
%%
%%Thus, for a conservative practitioner,
%%the E-FDS methods appear more appealing
%%once a sufficiently accurate bias potential is available.
%%
%To minimize this effect,
%many adaptive methods adopt a schedule to
%reduce the updating magnitude
%over the course of simulation\cite{
%marsili2006,
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%barducci2008}.
%%
%%Thus, in late stages of a long simulation,
%%the updating becomes negligible and
%%the adaptive simulation tends to
%%the corresponding equilibrium one.
%%
%%if the bias potential were converged to the exact value,
%%this adaptive FDS tends to the ideal equilibrium counterpart,
%%particular equilibrium simulation can be thought as
%%the optimal asymptotic limit of the adaptive counterpart,
%%
%For example, one can show that
%if the magnitude is controlled by
%the optimal inverse-time prescription\cite{
%belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
%liang2007,
%morozov2007, zhou2008,
%komura2012, *caparica2012, *caparica2014},
%the WL algorithm is equivalent to the ideal equilibrium counterpart
%in improving the precision of the bias potential,
%thus allowing for maximum efficiency
%(cf. Appendix \ref{sec:equilerr}).
%%the bias potential in the WL algorithm
%%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%%with the precision of the histogram as simulation lengthens.

@@

%One can show that with a constant magnitude,
%$\alpha(t) = \alpha > 0$,
%the distribution collected from
%an adaptive FDS simulation
%is identical to $p_i$ in the long time or asymptotic limit.
%
%However, as the above updates
%change $V_i(t)$ continuously,
%the equilibrium distribution,
%Eq. \eqref{eq:pi_p_V}, no longer holds
%at all times,
While adaptive FDS simulations
can often sample the desired distribution, $p_i$,
even with a constant magnitude,
$\alpha(t) = \alpha$,
frequent updates to the bias potential
disrupt the underlying equilibrium sampling
and introduce artificial errors that need to be
removed by gradually decreasing the updating magnitude\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, zhou2008,
  laio2005, bussi2006, poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}.
%%
%The source of the deviation is two-fold.
%%
%First, since $V_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%

@@

%For the single-bin scheme
%used in WL algorithm,
A more effective way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
An interesting explanation of this result\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average computed from all data points collected so far,
so that the addition of a new data point only
affects the average by its due weight, $1/t$,
which is the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
Alternatively, we may derive it
from optimally planning the simulation lengths in the stages
and the reduction factor in the WL setup (cf. Appendix \ref{sec:wltoinvt}).

%In the next section,
%we will re-derive this formula
%and suggest a slight modification
%whereby the origin of $t$ is shifted
%according to the initial error.
%

%As Eq. \eqref{eq:alpha_invt} was being developed for the WL algorithm,
%several similar ways of
%reducing the updating magnitude
%were developed for umbrella sampling and metadynamics,
%particularly, the self-healing umbrella sampling
%and well-tempered metadynamics\cite{
%  marsili2006, barducci2008, dickson2011, dama2014}.
%%
%In this study, we wish to find the optimal schedule, $\alpha(t)$,
%for a general updating scheme,
%including the Gaussian scheme used in metadynamics.

@@

In this section,
we will derive the optimal $\alpha(t)$
for the single-bin scheme,
Eq. \eqref{eq:wl_update},
as a preparation
for the more general multiple-bin scheme.
%
We will suggest a slight modification of Eq. \eqref{eq:alpha_invt}
whereby the origin of $t$ is shifted by an amount
determined from the initial error.
%
We will first express the error of
the bias potential, $V_i(t)$,
as a functional of $\alpha(t)$,
and then minimize it by variation.
%

\subsubsection{Error definition}

First, we define the net error in the bias potential
by deducting two contributions.
%
The first contribution comes from the target value,
Eq. \eqref{eq:Vi_target}.
%
For this purpose,
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  V_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i$'s are
equivalent to the updates to $V_i$'s,
since the shift
remains a constant during the course of simulation,
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V},

@@

Note that the error weighting distribution, $\pmb\rho$,
defined above
is independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$.
%
In practice, $\pmb\rho$ is usually flat,
i.e. $\rho_i \equiv 1/n$,
even when $\mathbf p$ is not
(e.g. in well-tempered metadynamics,
$\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
  barducci2008, dama2014}:
$\ln p_i = \ln {p^*_i}/(1+\Delta) + \mathrm{const.}$ for some positive $\Delta$).
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%

@@

%
For the random noise part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
and we will assume it is white, i.e.
%The noise is not necessarily white,
%and the autocorrelation function
%depend only on the time difference
%
\begin{equation}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  =
  %\sigma_{ij}(t - t')
  G^{(0)}_{ij} \, \delta(t - t')
  ,
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
for some value of $G^{(0)}_{ij}$,
and $\delta(t)$ is the Dirac's delta function.
%where $\sigma_{ij}(t)$ is an even function
%that vanishes at large $t$.
%
This approximation is valid if the simulation period, $T$,
is much longer than the autocorrelation time of noise.
%
For non-white noise,
the equivalent value of $G^{(0)}_i$ can be approximately obtained
from the integral,
$\int_{-\infty}^\infty \sigma_{ij}(t) \, dt$,
or in the discrete time case,
$\sum_{-\infty}^\infty \sigma_{ij}(t)$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
%
where
%
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also governed by white noise as is $\zeta_i(t)$,
i.e. we may write
$$
  \langle \xi_i(t) \, \xi_i(t') \rangle = G_i \, \delta(t - t'),
$$
for some $G_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

The total error
at the end of the simulation, $t = T$,
as defined in Eq. \eqref{eq:error_sum},
now depends implicitly on the function $\alpha(t)$
via $x_i(T)$.
%
Our aim is to find the $\alpha(t)$
that minimizes this error.
%%
%It suffices to consider
%a one-variable problem that
%minimizes each individual
%$\left\langle x_i^2(T) \right\rangle$.

%\subsubsection{\label{sec:onevar}
%Optimization of a one-variable problem}
